Efficient FPGA Implementation of\\ Digit Parallel Online Arithmetic Operators

### 2014-06-30 FPT2014/bare_conf.tex
Georges comment 3_1

\begin{abstract}
%\boldmath
Online arithmetic has been widely studied for ASIC implementation. Online components were originally designed to perform computations most significant digit first, resulting in the ability to chain arithmetic operators together for low latency. More recently, research has shown that digit parallel online operators can fail more gracefully when operating beyond the deterministic clocking region in comparison to operators with conventional arithmetic. Unfortunately, the utilization of online arithmetic operators in the past has required a large area overhead for FPGA implementation. In this paper, we propose novel approaches to implement the key primitives of online arithmetic, adders and multipliers, efficiently on modern Xilinx FPGAs with 6-input LUTs and carry resources. We demonstrate experimentally that in comparison to a direct implementation, the proposed architectures achieve slice savings of over $67\%$ and $69\%$, and speed-ups of over $1.2\times$ and $1.5\times$ for adders and multipliers, respectively. Furthermore in comparison to standard arithmetic primitives, this translates to reduction of the area overheads of up to $8.41\times$ and $8.11\times$ for adders and multipliers using direct implementation, to $1.88\times$ and $0.54\times$ using our approach.

@@

Recent research has demonstrated another key advantage of online arithmetic: that it allows graceful degradation when operating beyond the deterministic clocking region, i.e. when overclocking \cite{SKDAC14_REVIEW}. This theme of research stems from the inspiration that for certain applications, releasing the absolute accuracy requirements could lead to significant performance improvements. However, traditional forms of computer arithmetic do not fail gracefully when timing violation happens, because timing errors tend to affect MSDs of the results. In comparison, online arithmetic is more ``overclocking friendly'', as timing errors only occur at LSDs of the results.

Despite its attractiveness, the use of online arithmetic on existing FPGAs is still limited due to a large area overhead. For standard arithmetic, because it is difficult to efficiently map arithmetic circuits solely onto look-up-tables (LUTs), both main commercial FPGA vendors introduced dedicated carry logic to reduce carry propagation delay~\cite{Virtex6}. Significant prior work explores methods that can effectively map circuits to the carry logic for performance improvements~\cite{FPL10_FPGA_CarryChain}, or proposes alternative carry logic for FPGAs~\cite{FPGA_CarryChain_New1,FPGA_CarryChain_New2}. Hard DSP blocks are also included in modern architectures to improve the area and performance of multiplications and multiply accumulates~\cite{XilinxDSP}. However, most of these approaches are designed to accelerate computations with the conventional form of arithmetic.

For online arithmetic, two major research directions have been investigated. Firstly, there has been some research into FPGA implementations of online adders; a brief review of relevant works in this area is presented in Section~\ref{Sec:OA_FPGA}. Secondly, there are existing approaches to build general purpose online adders specifically for FPGAs with 4-input LUTs and with two LUTs within a slice or a logic element (LE), such as the Xilinx Spartan~3 series \cite{XilinxSpartan} and the Altera Cyclone III \cite{AlteraCyclone}. However, as will be discussed in Section~\ref{Sec:OA_FPGA}, these approaches cannot be directly applied on many modern FPGAs, which typically have 6-input LUTs and four LUTs in a slice, such as the Xilinx Virtex series FPGAs and all Xilinx 7 series FPGAs~\cite{Virtex7}. Moreover, the move to the 6-LUT on the basic architecture opens new opportunities for further optimization.
%However, the implementations of other key arithmetic primitives, such as online multipliers, lack exploration.

In this paper, both issues will be addressed. In Section~\ref{Sec:OA_FPGA}, we propose a novel approach to map digit parallel online adders onto FPGAs with 6-LUTs. The mapping is based on the fast carry logic and ensures the available logic resources within a slice are fully utilized. In Section~\ref{Sec:OM_FPGA}, we optimize the online multiplication algorithm to yield an efficient digit parallel online multiplier for FPGAs. We demonstrate experimentally on Xilinx Virtex-6 FPGAs that the proposed designs achieve significant area reduction and performance gain over their original implementations. For online adders, our method achieves  slice savings of over $67\%$ and  frequency speed-ups of over $1.2\times$ in comparison to a direct implementation. In comparison to the standard ripple-carry adder, the area overhead drops from up to $8.41\times$ to $1.88\times$ using the proposed online adder. For online multipliers, the slice utilization of the proposed architecture drops of over $69\%$, with a frequency speed-up over $1.5\times$ in comparison to a direct implementation. When compared to the multiplier with standard arithmetic, the direct implementation has an area overhead up to $8.11\times$, which drops to $1.84\times$ when using the proposed new architecture.

Our approach has the additional advantage of being able to create a correctly rounded multiplier with a reduced output precision using less hardware than a standard multiplier, which multiplies two $N$-digit numbers and produces an output of $2N$ digits. However, in many cases the full precision product is not required. In such cases, a reduced precision output is created either by truncating or rounding the LSDs. Although there has been research into saving area by avoiding the creation of unnecessary LSDs~\cite{TheoTC}, these techniques still require extensive simulation to guarantee correctness. In contract, since online multipliers compute the output from the MSD first, hardware can be saved by not generating the LSDs in the first place. As an example, if we are only interested in the most significant half of the product, we can create a digit parallel online multiplier that saves upto $56\%$ slice usage in comparison to a standard multiplier with truncated output.

>>>>

\begin{abstract}
%\boldmath
Online arithmetic has been widely studied for ASIC implementation. Online components were originally designed to perform computations in digit serial with most significant digit first, resulting in the ability to chain arithmetic operators together for low latency. More recently, research has shown that digit parallel online operators can fail more gracefully when operating beyond the deterministic clocking region in comparison to operators with conventional arithmetic. Unfortunately, the utilization of online arithmetic operators in the past has required a large area overhead for FPGA implementation. In this paper, we propose novel approaches to implement the key primitives of online arithmetic, adders and multipliers, efficiently on modern Xilinx FPGAs with 6-input LUTs and carry resources. We demonstrate experimentally that in comparison to a direct RTL synthesis, the proposed architectures achieve slice savings of over $67\%$ and $69\%$, and speed-ups of over $1.2\times$ and $1.5\times$ for adders and multipliers, respectively. As a result, the area overheads of using online adders and multipliers in place of traditional arithmtic primitives is reduced from 
% Furthermore in comparison to standard arithmetic primitives, this translates to reduction of the area overheads from up to $8.41\times$ to $1.88\times$ for adders, and from $8.11\times$ to $1.84\times$ for multipliers using direct implementation, to $1.88\times$ and $0.54\times$ using our approaches.

@@

Recent research has demonstrated another key advantage of online arithmetic: that it allows graceful degradation when operating beyond the deterministic clocking region,  \textit{i.e.} when overclocking \cite{SKDAC14_REVIEW}. This theme of research stems from the inspiration that for certain applications, releasing the absolute accuracy requirements could lead to significant performance improvements. However, traditional forms of computer arithmetic do not fail gracefully when timing violation happens, because timing errors tend to affect MSDs of the results. In comparison, online arithmetic is more ``overclocking friendly'', as timing errors only occur at LSDs of the results.

Despite its attractiveness, the use of online arithmetic on existing FPGAs is still limited due to a large area overhead. For standard arithmetic, because it is inefficient and slow to map arithmetic circuits solely onto look-up-tables (LUTs), both main commercial FPGA vendors have introduced dedicated carry logic~\cite{Virtex6}. Significant prior work explores methods that can effectively map circuits to the carry logic for performance improvements~\cite{FPL10_FPGA_CarryChain}, or proposes alternative carry logic for FPGAs~\cite{FPGA_CarryChain_New1,FPGA_CarryChain_New2}. Hard DSP blocks are also included in modern architectures to improve the area and performance of multiplications and multiply accumulates~\cite{XilinxDSP}. However, most of these approaches are designed to accelerate computations with the conventional form of arithmetic.

For online arithmetic, two major research directions have been investigated. Firstly, there has been some research into FPGA implementations of online adders; a brief review of relevant work in this area is presented in Section~\ref{Sec:OA_FPGA}. Secondly, there are existing approaches to build general purpose online adders specifically for FPGAs with 4-input LUTs and with two LUTs within a slice or a logic element (LE), such as the Xilinx Spartan~3 series \cite{XilinxSpartan} and the Altera Cyclone III \cite{AlteraCyclone}. However, as will be discussed in Section~\ref{Sec:OA_FPGA}, these approaches cannot be directly applied on many modern FPGAs, which typically have 6-input LUTs and four LUTs in a slice, such as the Xilinx Virtex series FPGAs and all Xilinx 7 series FPGAs~\cite{Virtex7}. Moreover, the move to the 6-LUT on the basic architecture opens new opportunities for further optimization.
%However, the implementations of other key arithmetic primitives, such as online multipliers, lack exploration.

In this paper, both issues will be addressed. In Section~\ref{Sec:OA_FPGA}, we propose a novel approach to map digit parallel online adders onto FPGAs with 6-LUTs. The mapping is based on the fast carry logic and ensures the available logic resources within a slice are fully utilized. In Section~\ref{Sec:OM_FPGA}, we optimize the online multiplication algorithm to yield an efficient digit parallel online multiplier for FPGAs. We demonstrate experimentally on Xilinx Virtex-6 FPGAs that the proposed designs achieve significant area reduction and performance gain over their original implementations. For online adders, our method achieves  slice savings over $67\%$ and  frequency speed-ups of over $1.2\times$ in comparison to a direct implementation. In comparison to the standard ripple-carry adder, the area overhead drops from up to $8.41\times$ to $1.88\times$ using the proposed online adder. For online multipliers, the slice utilization of the proposed architecture drops of over $69\%$, with a frequency speed-up over $1.5\times$ in comparison to a direct implementation. When compared to the multiplier with standard arithmetic, the direct implementation has an area overhead up to $8.11\times$, which drops to $1.84\times$ when using the proposed new architecture.

Our approach has the additional advantage of being able to create a correctly rounded multiplier with a reduced output precision using less hardware than a full multiplier, which multiplies two $N$-digit numbers and produces an output of $2N$ digits. However, in many cases the full precision product is not required. In such cases, a reduced precision output is created either by truncating or rounding the LSDs. Although there has been research into saving area by avoiding the creation of unnecessary LSDs~\cite{TheoTC}, these techniques still require extensive simulation to guarantee correctness. In contract, since online multipliers compute the output from the MSD first, hardware can be saved by not generating the LSDs in the first place. As an example, if we are only interested in the most significant half of the product, we can create a digit parallel online multiplier that saves up to $56\%$ slice usage in comparison to a standard multiplier with truncated output.
